---
title: "Peer Assessment"
author: "Ricardo Alcaraz Fraga"
date: "1/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document is made as requested by Roger D. Peng as part of the Coursera
course **Reproducible Research**. I'm going to explain the procedure I followed
in order to answer to the assessment's questions.  
  
## Reading the data
As a way of organizing the project files, I designed a folder named **Data** in
which I always store the data that is going to be used in the project.

``` {r reading data}
data <- read.csv('/home/ricardo/Escritorio/PeerAssessment/Data/activity.csv')
head(data)
```

## What is mean total number of steps taken per day?
The first question we are going to answer in this question is: *how many steps
were taken by day?* This can be seen in the next histogram.

``` {r total steps,echo = FALSE, fig.heigth = 2, fig.width = 10}
library(ggplot2)
dates <- unique(data[, 'date'])

# Computing mean by day
sums <- c()
for(date_i in dates) {
      value_day <- subset(data, as.character(date) == 
                                as.character(date_i))$steps
      
      sums <- c(sums, sum(value_day, na.rm = TRUE))
}

frame <- data.frame(date = as.Date(dates, '%Y-%m-%d'), steps = sums)

print(ggplot(frame, aes(x = date, y = steps)) + 
            geom_bar(stat = 'identity', color = '#363A19', fill = '#363A19', 
                     alpha = 0.8))
```

There were days in which no data was registered for the number of steps, these 
were treated as if no steps were taken that day.  
The second question answered in this section is *what is the mean and median of
the numbers of steps taken each day?*

``` {r mean and median by day, echo = FALSE}
# Computing mean by day
means <- c()
medians <- c()

for(date_i in dates) {
      value_day <- subset(data, as.character(date) == 
                                as.character(date_i))$steps
      
      mean_day <- mean(value_day, na.rm = TRUE)
      
      if(is.na(mean_day))
        mean_day = 0
      
      median_day <- median(value_day, na.rm = TRUE)
      
      means <- c(means, mean_day)
      medians <- c(medians, median_day)
}

frame <- data.frame(date = as.Date(dates, '%Y-%m-%d'),
                    mean = means, median = medians)
print(frame)
```

As you can see, the median value is not that helpful because of all the NAs and 
0s on the data, still, the mean is indeed helpful and this numbers compliment 
really good the histogram previously shown.  

## What is the average daily activity pattern?
In order to observe this question I'm presenting a time series plot of the 5 
minute interval and the average numbers of steps taken averaged across all days.  

``` {r mean steps by day, echo = FALSE, fig.heigth = 2, fig.width = 10}
print(ggplot(frame, aes(x = date, y = mean)) +
        geom_line(color = 'purple'))
```

Continuing with this section I'll show what's the interval with the greatest 
mean of steps.

``` {r greatest mean, echo = FALSE}
intervals <- unique(data$interval)

values_intervals <- c()
for(interval_i in intervals) {
   steps_i <- subset(data, interval == interval_i)$steps
   mean_steps_i <- mean(steps_i, na.rm = TRUE)
   
   values_intervals <- c(values_intervals, mean_steps_i)
}

frame <- data.frame(interval = intervals, mean = values_intervals)
print(frame[which.max(frame$mean), ])
```

## Imputing missing values
As we've seen, there are a lot of NAs values that may bias our measurements 
means and medians. Let's look at how many NA values are there in the data set. 
``` {r NAs}
len <- length(data$steps)
print(paste0('Total of observations in the data set: ', len))

nas <- sum(is.na(data$steps))
print(paste0('Total of NA values in the data set: ', nas))

perc <- nas * 100 / len
print(paste0(paste0('Percentage of NA values: ', perc), '%'))
```

To fill in the NA values, we were told to create a strategy for imputing this 
values. We are going to implement a strategy in which we fill the gaps with the 
mean for that 5-minute interval.

``` {r imputing}
nas_ind <- which(is.na(data$steps))

for(index in nas_ind){
  row_index_frame <- which(frame$interval == data[index, 'interval'])
   data$steps[index] <- frame[row_index_frame, 'mean']
}

print(head(data))
```

As we can see from the head from the data, there are now no NA values! Now let's
re-compute the histogram of the total of steps taken each day, as well as the 
mean and median and we'll see how different it looks.  

``` {r imputed histogram, echo = FALSE}
# Computing mean by day
sums <- c()
for(date_i in dates) {
      value_day <- subset(data, as.character(date) == 
                                as.character(date_i))$steps
      
      sums <- c(sums, sum(value_day, na.rm = TRUE))
}

frame <- data.frame(date = as.Date(dates, '%Y-%m-%d'), steps = sums)

print(ggplot(frame, aes(x = date, y = steps)) + 
            geom_bar(stat = 'identity', color = '#363A19', fill = '#363A19', 
                     alpha = 0.8))

# Computing mean by day
means <- c()
medians <- c()

for(date_i in dates) {
      value_day <- subset(data, as.character(date) == 
                                as.character(date_i))$steps
      
      mean_day <- mean(value_day, na.rm = TRUE)
      
      if(is.na(mean_day))
        mean_day = 0
      
      median_day <- median(value_day, na.rm = TRUE)
      
      means <- c(means, mean_day)
      medians <- c(medians, median_day)
}

frame <- data.frame(date = as.Date(dates, '%Y-%m-%d'),
                    mean = means, median = medians)
print(frame)
```

What a change! Everything looks a lot useful now (of course, if we assume our 
imputing strategy is good enough to generate better insight from the data).  

## Are there differences in activity patterns between weekdays and weekends?

``` {r week days, echo = FALSE, fig.heigth = 2, fig.width = 10}
days <- c()
for(i in 1:length(frame$mean)) {
  
  if(weekdays(frame[i, 'date']) %in% c('lunes', 'martes', 'miércoles', 
                                       'jueves', 'viernes')) {
    days <- c(days, 'weekday')
  }
  
  else
    days <- c(days, 'weekend')
}

frame <- data.frame(date = frame$date,
                    mean = frame$mean, 
                    median = frame$median,
                    day = days)

print(ggplot(frame, aes(x = date, y = mean, group = day, color = day)) + 
        geom_line() + 
        facet_wrap(~day))

```
